Spatial Reasoning and Data Displays

### 2015-01-07 Paper/Draft1/TestingVisualAptitude.Rnw
Paragraph linking lineups to graphical displays and justifying spatial reasoning tests.

\IEEEtitleabstractindextext{%
\begin{abstract}
Graphics convey statistical information very efficiently, but rely on a different set of mental processes than tabular displays. This study examines the demographic characteristics and visual skills associated with perception of statistical lineups. We conclude that statistical lineups are essentially a classification test in a visual domain, and that performance on the lineup protocol is associated with general aptitude, rather than specific tasks such as card rotation and spatial manipulation. We also examine the possibility that specific graphical tasks may be associated with certain visual skills and conclude that more research is necessary to understand which visual skills are required in order to understand certain plot types. 
\end{abstract}

@@

% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps (small caps for compsoc).
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{S}{tatistical} graphics provide quick summaries of data, models, and results, but these displays may not be equally useful to all viewers. Mathematical ability is important, even for the simplest graphs:  mathematical ability, not spatial ability, was shown~\cite{shah1995conceptual} to be associated with accuracy on a simple two-dimensional line graph.

Spatial ability becomes more important, however, when more complicated graphical displays are used in comparison tasks: the lower performance of individuals with low spatial ability on tests utilizing diagrams and graphs is attributed ~\cite{mayer1994whom} to the fact that more cognitive resources are required to process the visual stimuli, which leaves fewer resources to make connections and draw conclusions from those stimuli. It is theorized that graphics are a form of ``external cognition"~\cite{scaife1996external} that guide, constrain, and facilitate cognitive behavior\cite{zhang1997nature}. 
Graphics utilize higher-bandwidth visual pathways to encode information\cite{baddeley1974working}, allowing viewers to quickly and intuitively encode multiple dimensions of numerical relationships; however, this higher-bandwidth approach also depends on the viewer's visual and spatial reasoning ability. Well-designed graphics emphasize and present important features of the data while minimizing  features of lesser importance, guiding the viewer towards conclusions that are meaningful in context and supported by the data while maximizing the information encoded in working memory. Under this framework, well-designed graphics reduce memory load and make more cognitive resources available for other tasks (such as drawing conclusions from the data), at the cost of depending on certain visuospatial reasoning abilities.

Many theories of graphical learning center around the difference between visual and verbal processing: the dual-coding theory emphasizes the utility of complementary information in both domains, while the visual argument hypothesis emphasizes that graphics are more efficient tools for providing data with spatial, temporal, or other implicit ordering, because the spatial dimension can be represented graphically in a more natural manner~\cite{vekiri2002value}. Both of these theories suggest spatial ability would impact a viewer's use of graphics, because spatial ability either influences cognitive resource allocation or affects the processing of spatial relationships between graphical elements. In addition, previous investigations into graphical learning and spatial ability have found relationships between spatial ability and the ability to read information from graphs~\cite{lowrie2007solving}.

The lineup protocol~\cite{hofmann2012graphical,wickham2010graphical, buja2009statistical} is a testing framework that allows us to quantify the significance of graphical findings with the same mathematical rigor as conventional hypothesis tests.

>>>>

\IEEEtitleabstractindextext{%
\begin{abstract}
\comment{Tone down the `statistical' part ... statistical, to the wrong ears sounds complicated. } Graphics convey numerical information very efficiently, but rely on a different set of mental processes than tabular displays. 
\newtext{}
This study examines the demographic characteristics and visual skills associated with perception of graphical lineups. We conclude that lineups are essentially a classification test in a visual domain, and that performance on the lineup protocol is associated with general aptitude, rather than specific tasks such as card rotation and spatial manipulation. We also examine the possibility that specific graphical tasks may be associated with certain visual skills and conclude that more research is necessary to understand which visual skills are required in order to understand certain plot types. 
\end{abstract}

@@

% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps (small caps for compsoc).
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{D}{ata} displays provide quick summaries of data, models, and results, but not all displays are equally good, nor is any data display equally useful to all viewers. 
Graphics utilize higher-bandwidth visual pathways to encode information~\cite{baddeley1974working}, allowing viewers to quickly and intuitively relate multiple dimensions of numerical quantities.
Well-designed graphics emphasize and present important features of the data while minimizing  features of lesser importance, guiding the viewer towards conclusions that are meaningful in context and supported by the data while maximizing the information encoded in working memory. Under this framework, well-designed graphics reduce memory load and make more cognitive resources available for other tasks (such as drawing conclusions from the data), at the cost of depending on certain visuospatial reasoning abilities.

Many theories of graphical learning center around the difference between visual and verbal processing: the dual-coding theory emphasizes the utility of complementary information in both domains, while the visual argument hypothesis emphasizes that graphics are more efficient tools for providing data with spatial, temporal, or other implicit ordering, because the spatial dimension can be represented graphically in a more natural manner~\cite{vekiri2002value}. Both of these theories suggest spatial ability would impact a viewer's use of graphics, because spatial ability either influences cognitive resource allocation or affects the processing of spatial relationships between graphical elements. In addition, previous investigations into graphical learning and spatial ability have found relationships between spatial ability and the ability to read information from graphs~\cite{lowrie2007solving}. 
However,  mathematical ability, not spatial ability, was shown~\cite{shah1995conceptual} to be associated with accuracy on a simple two-dimensional line graph. 
Spatial ability becomes more important when more complicated graphical displays are used in comparison tasks: the lower performance of individuals with low spatial ability on tests utilizing diagrams and graphs is attributed ~\cite{mayer1994whom} to the fact that more cognitive resources are required to process the visual stimuli, which leaves fewer resources to make connections and draw conclusions from those stimuli. It is theorized that graphics are a form of ``external cognition"~\cite{scaife1996external} that guide, constrain, and facilitate cognitive behavior\cite{zhang1997nature}.

\newtext{
"Lineups" are often used to evaluate the effectiveness of different graphical displays \cite{kosara2010mechanical,wickham2010graphical,majumder2013validation}; as the police counterpart, lineups consist of several distractor plots and one target. 
This method is described in more detail in the next section; here we motivate the importance of understanding the connection between lineups and spatial reasoning. 
}\newtext{
Lineups provide a quantitative measurement of the effectiveness of a particular plot: if participants consistently identify the target plot rather than the randomly-generated distractors, the plot effectively shows the difference between real data and random noise. 
This removes some of the subjectivity from user evaluations of display effectiveness, and the procedure is simple enough that it does not generally require familiarity with data-based graphics. 
}\newtext{
While previous research has examined the link between certain types of graphical perception and spatial skills, it is important to identify any additional visual skills required by the lineup task, as well as better understand demographic characteristics (math education, research experience, age, gender) which may differentially impact performance\cite{humanfactorslineups}. 
Exploring the participant characteristics which contribute to successful lineup evaluation allows us to better understand the lineup task and potential sources of variability in the data. 
}

% \comment{Here is a paragraph missing, that connects the existing literature to what we are doing in this paper. 
% In particular, we need to answer the questions: 
% 
% - why are we interested in evaluating which skills are necessary to evaluate graphical displays (this has been implicitly answered by saying that cognitive load was reduced, so we free up resources and make graphics a better tool for more people.
% 
% - what are the limits of the studies that you have cited, ie is there something that we can do better?
% 
% - what advantage do lineups bring to the table?
% }
% 
% 
% \comment{How do perceptual principles apply to statistical graphics (data displays)? - Lineups help us to isolate different parts of displays and allow in-built quantification of results.}
% 
% 
% \comment{Why are lineups variable? - Individual variability is big, we want to investigate which abilities individuals help the most in evaluating. - for this reason we need to motivate why lineups are useful. }
% 
% 
% \comment{We examine 'some demographic characteristics' is a bit lukewarm. I think once the missing paragraph is there, you can sell the rest of the paper a bit more passionately.}

\section{Methods}

\subsection{The Lineup Protocol}
The lineup protocol~\cite{hofmann2012graphical,wickham2010graphical, buja2009statistical} is a testing framework that allows us to quantify the significance of graphical findings with the same mathematical rigor as conventional hypothesis tests.
